---
title: "How to use flscuts miloR wrappers"
output: html_document
date: "2024-09-23"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rm(list=ls())
# Define dataset paths depending on your machine
if (grepl("^gizmo", Sys.info()["nodename"])) {
  ROOT_DIR1 <- "/fh/fast/furlan_s/user/sbhise"
  stem <- "Pitt_2"
} else {
  ROOT_DIR1 <- "/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments"
  stem <- "Pitt/Pitt_2"
}


DATA_DIR <- file.path(ROOT_DIR1,  stem, "data")      # SPECIFY HERE
RES_DIR  <- file.path(ROOT_DIR1, stem, "res")     # SPECIFY HERE
RMD_DIR  <- file.path(ROOT_DIR1, stem, "rmd")     # SPECIFY HERE
CDS_DIR <- file.path(ROOT_DIR1,  stem, "cds/final")
FIG_DIR <- file.path(ROOT_DIR1,  stem, "figs")

```


## **Installing flscuts**

First, ensure you have the `devtools` R package installed, which allows you to install packages from GitHub. If `devtools` is installed, you can easily install using the following command:

```{r, eval=F}
devtools::install_github("furlan-lab/flscuts")
```

## **Loading data**

In this section, we’ll load a Seurat object and create some colors:  
- **Dataset (`seu`)**: Contains the data you want to run miloR on.  


```{r}
# Load required packages
suppressPackageStartupMessages({
  library(flscuts)
  library(viewmastR)
    library(Seurat)
    library(scCustomize)
    library(magrittr)
    library(miloR)
    library(scater)
  library(patchwork)
})

set.seed(1234) # Set Seed
#devtools::install_github("broadinstitute/infercnv", force=TRUE)

group_cols<-c("Pre_Ifng_1_BM" = "#594286", 
                "Post_Ifng_1_BM" = "#bd9895", 
                "Pre_Ifng_1_34" = "#f76b14", 
                "Post_Ifng_1_34" = "#20B2AA", 
                "Pre_Ifng_5_BM" = "orange", 
                "Post_Ifng_5_BM" = "grey", 
                "Pre_Ifng_7_BM" = "violetred4", 
                "Post_Ifng_7_BM" = "lightgreen")

seu <- readRDS(file.path(CDS_DIR, "Patient_1_final_CDS.RDS"))
seu <- seu[,seu$celltype %in% c("01_HSC", "04_Myeloid_Progenitor",  "02_Early_Erythroid")]
#downsample
seu <- seu[,sample(1:dim(seu)[2], 2000)]
#DimPlot(seu, group.by = "geno")
seu <- seu[,seu$geno %in% "1"]
seu$rx <- strsplit(colnames(seu), "_") %>% sapply("[[", 1)
#DimPlot(seu, group.by="rx")
#DimPlot(seu, group.by = "celltype")
```

Next we run the make_nhoods function.  This turns a seurat object into a SingleCellExperiment object and does the necessary calculation of neighborhoods.  Make sure you have the default assay set to "RNA"
Also make sure your factor levels are "clean".  In the code above we removed all celltypes except the 3 listed.  You will note that doing so doesn't reset the factor levels, so make sure to refactor.  Also make sure your variables are all factors

```{r}
#seu$celltype<- factor(as.character(seu$celltype))

#roxygen2::roxygenise()
DefaultAssay(seu)<-"RNA"

seud <- make_dummy_variable(seu, joining_variable = "rx")
scm <- make_nhoods(seud, variable="rx", sample = "sample")
da_result <- test_nhoods(scm, "~rx")

ggplot(da_result, aes(PValue)) + geom_histogram(bins=50)

umap_pl1 <- plotReducedDim(scm$milo, dimred = "UMAP", colour_by="rx") +
  guides(fill="none")+theme_void()
umap_pl2 <- plotReducedDim(scm$milo, dimred = "UMAP", colour_by="celltype") +
  guides(fill="none")+theme_void()+scale_color_manual(values=sfc(10))
nh_graph_pl <- plotNhoodGraphDA(scm$milo, da_result, layout="UMAP",alpha=0.05)

umap_pl1 + umap_pl2 + nh_graph_pl +
  plot_layout(guides="collect")
```

## **Appendix**

```{r}
sessionInfo()
getwd()
```
