---
title: "How to make flow style plots"
output: html_document
date: "2024-10-18"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rm(list=ls())
# Define dataset paths depending on your machine
if(grepl("^gizmo", Sys.info()["nodename"])){
  ROOT_DIR1<-"/fh/fast/furlan_s/experiments/patient_marrows/aggr/cds/indy"
  ROOT_DIR2<-"//fh/fast/furlan_s/experiments/patient_marrows"
} else {
  ROOT_DIR1<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/patient_marrows/aggr/cds/indy"
  ROOT_DIR2<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/patient_marrows/5336/cds"
}
knitr::opts_chunk$set(message= FALSE, warning = FALSE)
```

## **Installing flscuts**

First, ensure you have the `devtools` R package installed, which allows you to install packages from GitHub. If `devtools` is installed, you can easily install using the following command:

```{r, eval=F}
devtools::install_github("furlan-lab/flscuts")
```

## **Loading data**

In this section, we’ll load two Seurat objects, fix the celltypes so they harmonize, and create some colors.

```{r}
# Load required packages
suppressPackageStartupMessages({
  library(flscuts)
  library(Seurat)
  library(viewmastR)
  library(ggplot2)
  library(scCustomize)
})
```


#### Load AML dataset and define the leukemia
```{r}
seu <- readRDS(file.path(ROOT_DIR1, "220831_NB1.RDS"))
DimPlot_scCustom(seu, group.by = "geno")
DimPlot_scCustom(seu, group.by = "seurat_clusters")
seu$celltype <-as.character(seu$celltype)
seu$celltype[seu$seurat_clusters %in% c(0,1,5,3,14)]<- "00_LEUMEKIA"
DimPlot_scCustom(seu, group.by = "celltype")
```

```{r}
DefaultAssay(seu)<-"ADT"
cts <- seu@assays$ADT@counts
rownames(cts) <- gsub("\\.1", "", rownames(cts))
rownames(cts)[140] <- "CD15"
seu$ADT <- CreateAssayObject(cts)
seu<-NormalizeData(seu, normalization.method = "CLR")
seu$adt_sum <- colSums(seu@assays$ADT$counts)
seu$log_adt_sum <- log10(seu$adt_sum)
aml_panel(seu, color = "celltype", cols = as.character(pals::polychrome(15)))
```

#### Load ALL dataset and define the leukemia
```{r}
seu <- readRDS(file.path(ROOT_DIR2, "KMGall"))
DimPlot_scCustom(seu, group.by = "geno")
DimPlot_scCustom(seu, group.by = "seurat_clusters")
seu$celltype <-as.character(seu$celltype)
seu$celltype[seu$seurat_clusters %in% c(0,1,5,3,14)]<- "00_LEUMEKIA"
DimPlot_scCustom(seu, group.by = "celltype")
```


## **Appendix**

```{r}
sessionInfo()
getwd()
```
