---
title: "How to use UMAP transform on a single cell dataset (Seurat) using Seurat Workflow"
output: html_document
date: "2024-09-26"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

Note that this code was inspired by and adapted from: 





```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rm(list=ls())
# Define dataset paths depending on your machine
if(grepl("^gizmo", Sys.info()["nodename"])){
  ROOT_DIR1<-"/fh/fast/furlan_s/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/fh/fast/furlan_s/grp/data/ddata/BM_data"
  ROOT_DIR3<-"/fh/fast/furlan_s/experiments/LRPBMC/cds"
} else {
  ROOT_DIR1<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf"
  ROOT_DIR3<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/LRPBMC/cds"
}
knitr::opts_chunk$set(message= FALSE, warning = FALSE)
```

## **Installing flscuts**

First, ensure you have the `devtools` R package installed, which allows you to install packages from GitHub. If `devtools` is installed, you can easily install using the following command:

```{r, eval=F}
devtools::install_github("furlan-lab/flscuts")
```

## **Loading data**

In this section, we’ll load two Seurat objects, fix the celltypes so they harmonize, and create some colors.

```{r}
# Load required packages
suppressPackageStartupMessages({
  library(flscuts)
    library(Seurat)
  library(monocle3)
  library(viewmastR)
  library(ggplot2)
  library(scCustomize)
  library(tidyr)
})
```


#### Load reference dataset, run Seurat workflow, and make some colors
```{r}
seur <- readRDS(file.path(ROOT_DIR1, "240926_final_object.RDS"))
# Make some colors
cols1 <- as.character(pals::polychrome()[c(1,3:13)])

seur <- NormalizeData(seur) %>% FindVariableFeatures() %>% ScaleData() %>% RunPCA() %>% RunUMAP(dims=1:30, return.model = T)
DimPlot_scCustom(seur, group.by = "celltype", colors_use = cols1)+theme_void()+ggtitle("Reference")
```

#### Load query dataset, run Seurat workflow
```{r}
seu <-readRDS(file.path(ROOT_DIR3, "240926_5p_seu.RDS"))
seu <- NormalizeData(seu) %>% FindVariableFeatures() %>% ScaleData() %>% RunPCA() %>% RunUMAP(dims=1:30, return.model = T)
DimPlot_scCustom(seu, group.by = "celltype", colors_use = cols1)+theme_void()+ggtitle("Query")
```


## **Make sure labels are correct**

In this code, we show that the labels given to the reference and query cells are correct.

#### First the reference
```{r, fig.width = 12, fig.height = 12}
Idents(seur) <- seur$celltype
all_markers <- FindAllMarkers(object = seur) %>%
    Add_Pct_Diff() %>% dplyr::filter(pct_diff > 0.6)
    

top_markers <- Extract_Top_Markers(marker_dataframe = all_markers, num_genes = 5, named_vector = FALSE,
    make_unique = TRUE)

Clustered_DotPlot(seurat_object = seur, features = top_markers, colors_use_idents = cols1, k=11, plot_km_elbow = F)
```


#### And the query
```{r, fig.width = 12, fig.height = 12}
Idents(seu) <- seu$celltype
all_markers <- FindAllMarkers(object = seu) %>%
    Add_Pct_Diff() %>% dplyr::filter(pct_diff > 0.6)
    

top_markers <- Extract_Top_Markers(marker_dataframe = all_markers, num_genes = 5, named_vector = FALSE,
    make_unique = TRUE)

Clustered_DotPlot(seurat_object = seur, features = top_markers, colors_use_idents = cols1, k=11, plot_km_elbow = F)
```



## Find Transfer Anchors

```{r, fig.height=10, fig.width=20}
anchors <- FindTransferAnchors(
    reference = seur,
    query = seu,
    reference.reduction = 'pca',
    features = rownames(x = seur[["pca"]]@feature.loadings),
    dims = 1:30,
    nn.method = "annoy",
    k.filter = NA,
    verbose = TRUE
  )

```


## Map using anchors
```{r, fig.height=8, fig.width=16}
seu.1 <- MapQuery(
    reference = seur,
    query = seu,
    anchorset = anchors,
    refdata = list(celltype = "celltype"),
    reference.reduction = "pca",
    reduction.model = "umap"
  )

p0 <- DimPlot_scCustom(seur, group.by = "celltype", label = T, colors_use = cols1) +theme_void()+ggtitle("Reference - aka Projector Dataset")
p1 <- DimPlot_scCustom(seu.1, reduction = "ref.umap",group.by = "predicted.celltype", label = T, colors_use = cols1) + NoLegend() +theme_void()+ggtitle("Query - Mapped using UMAP Projection from Reference")
p0+p1

```



## Alter parameters of UMAP model and redo projection

If you would like to alter the umap parameters you may do so as follows, then redo the projection.
```{r, fig.height=6, fig.width=16}
# set embeddings
seur[["umap.new"]] <- CreateDimReducObject(embeddings = seur[["umap"]]@cell.embeddings, key = "UMAPnew_", assay = "RNA")

# set UMAP models
umap.new.model <- list()
umap.new.model$n_epochs <- 1000
umap.new.model$alpha <-1
umap.new.model$method <- "umap"
umap.new.model$negative_sample_rate <- 5
umap.new.model$gamma <- 1
umap.new.model$approx_pow <- 0
umap.new.model$n_neighbors <- 10
umap.new.model$metric$cosine <- list()
umap.new.model$embedding <- seur[["umap.new"]]@cell.embeddings
ab_param <- uwot:::find_ab_params(spread = 1, min_dist = 0.3)
umap.new.model$a <- ab_param["a"]
umap.new.model$b <- ab_param["b"]
seur[["umap.new"]]@misc$model <- umap.new.model
 


seu.2 <- MapQuery(
  reference = seur,
  query = seu,
  anchorset = anchors,
  refdata = list(celltype = "celltype"),
  reference.reduction = "pca",
  reduction.model = "umap.new"
)


p2 <- DimPlot_scCustom(seu.2, reduction = "ref.umap",group.by = "predicted.celltype", label = T, colors_use = cols1) + NoLegend() +theme_void()+ggtitle("Query - Mapped using Altered UMAP Projection from Reference")

p0+p2



```


## **Appendix**

```{r}
sessionInfo()
getwd()
```
